---
title: "EFS Survival Analysis \n COG meeting Spring 2017 "
author: "Jenny Smith"
date: "March 29, 2017"
output: pdf_document
---

Purpose: To create EFS curves to identify fusions or mutations which are associated with patient outcome and provide evidence for re-stratification of risk classification. 

```{r setup}
library(knitr)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=50),tidy=TRUE, fig.align='center')
knitr::opts_knit$set(root.dir = 'H:/COG_Meetings_Analysis/2017.03.27_COG_spring2017/')
```


```{r message = FALSE, warning=FALSE}
library(survival)
library(RColorBrewer)
library(colorspace)
library(stringr)
library(Biobase)
library(dplyr)
library(magrittr)
library(tables)
getwd()
```


```{r}
source("H:/scripts/Survival_Analysis/Survplot_Functions.r")
```


#Read in the clinical data with the new classification strategy
```{r comment="", tidy=TRUE, size="tiny"}
# newClassification <- read.csv("New_stratification_outcomes_updated)_(003).csv")

newClassification <- read.csv("New_stratification_outcomes_updated_2017.03.28.csv")

head(newClassification[,c(1,9:12)])
```

```{r}
dim(newClassification)
```

```{r}
levels(newClassification$First.Event)
```

```{r}
newClassification$First.Event <- ifelse(newClassification$First.Event == "Censored", 0, 1)
```

```{r}
head(newClassification[,c(1,9:12)])
```


#Define Functions for subsetting and KM curves


##Subsetting for each fusion or mutation 
```{r}
NewClassSubsets <- function(df, type, string){
  if (type == "fusion"){
    clinDataSubset <- subset(df, grepl(string, newClassification$Reason))
  }
  
  if (type == "mutation"){
    clinDataSubset <- subset(df, grepl(string, newClassification$Gene.Mutation.Group))
  }
  return(clinDataSubset)
}
```


##Singleline KM survival estimates
```{r}
survivalAnalysis <- function(df){
    require(survival)
  
    #Event Free survival analysis
    efs <- survfit(Surv(time = (df$Event.Free.Survival.Time.in.Days/365), event = df$First.Event) ~ 1)
    
    return(efs)
}
```


##Customize the output of SurvivalPlot() function. 
SurvivalPlot is a function that is my generic version in Survplot_Functions.r
```{r}
efsCustomPlot <- function(fit, color){
    
    #label with number of patients
    if (length(fit$n) == 1){
      Num <- paste("n =", fit$n, sep = " ")
    }else{
      obs <- fit$n[1] + fit$n[2]
      Num <- paste("n =", obs, sep = " ")
    }
    
    #X axis position for Num
    pos <- (max(fit$time)*0.75)

    #customized plot 
    if (length(color > 1)){
      efs_plot <- SurvivalPlot(fit, "Mutation Status", "Years",color)
    }else{
      efs_plot <- SurvivalPlot(fit, "NA", "Years",color)
    }

    efs_plot <- efs_plot + annotate(geom="text", x=pos, y=1.0, label= Num, size=6)
    
    #return the ggplot object
    return(efs_plot)
}
```


Note: efs_plot + geom_label(aes(x=7, y=1.0, label=Num), size=5) would not recognize the Num variable and would cause error "object Num not found"

Solution: Create a second dataframe for geom_label with x,y coordinates and labels. 

http://stackoverflow.com/questions/30269619/r-ggplot-geom-txt-label-not-recognize-a-variable-in-function-call


#Subset the clinical Data for the family of fusions

###Fusions of interest per email communication
email on march 26, 2017 @ 11:00pm
```{r}
fusionNames <- c("ETS", "NUP98-KDM5A", "NUP98-NSD1", "RAM", "CBFA2T3-GLIS2", "NUP98")
MutationNames <- c("cohesin", "Spliceosome", "myosin")
```


###Loop through the fusions of interest and subset the clincal data. 
```{r}
fusionDataframes <- list()
for (fusion in fusionNames){
  fusionClinData <- NewClassSubsets(newClassification, "fusion", fusion) #subset the clinical data
  fusion <- gsub("-", "_", fusion) #change dash to underscore.
  fusionDataframes[[fusion]] <- fusionClinData #add dataframe to the list. 
}

# fusionDataframes
```

```{r}
#names of the data frames
summary(fusionDataframes)
```

```{r}
#subset nup98 fusions
NUP98_allOthers <- fusionDataframes[["NUP98"]][which( ! grepl("NUP98-KDM5A", fusionDataframes[["NUP98"]]$Reason) & ! grepl("NUP98-NSD1", fusionDataframes[["NUP98"]]$Reason )), ]
```

```{r}
#add to the list of dataframes
fusionDataframes[["NUP98_allOthers"]] <- NUP98_allOthers
```

```{r}
#Number of patients in each dataframe. 
sapply(fusionDataframes, nrow)
```

###Loop through the mutations of interest
```{r}
mutationsDataFrames <- list()
for (mut in MutationNames){
  mutationClinData <- NewClassSubsets(newClassification, "mutation", mut)
  mutationsDataFrames[[mut]] <- mutationClinData
}
```

```{r}
summary(mutationsDataFrames)
```

```{r}
sapply(mutationsDataFrames, nrow)
```




#Kaplan-Meier Curves of EFS (Event Free Survival)


##Results of Survival Analysis 
```{r}
#survival analysis (SA)
fusions_SA <- lapply(fusionDataframes, survivalAnalysis)

fusions_SA
```

```{r}
mutations_SA <- lapply(mutationsDataFrames, survivalAnalysis)

mutations_SA
```


##Plot the Survival Curves 

#### Event Free Survival 
```{r}
colorsFus <- c("dodgerblue4", "darkturquoise", "blueviolet", "firebrick", "deeppink4", "lightsteelblue4", "darkgreen")
colorsMut <- c("royalblue1","goldenrod", "springgreen3")
```

```{r message=FALSE, warning=FALSE}
efs_plots <- mapply(efsCustomPlot, fusions_SA, colorsFus, SIMPLIFY = FALSE)
```

```{r}
summary(efs_plots)
```

#### Loop to add the main titles
```{r}
for (fusion in names(efs_plots)){
  if (fusion == "RAM" ){
    title <- paste(c("Event Free Survival in \n Patients with", fusion, "Immunophenotype" ), collapse = " ")
  }else{
    title <- paste(c("Event Free Survival in \n Patients with", gsub("_", "-", fusion), " Fusions" ), collapse = " ")
  }
  print(title)
  efs_plots[[fusion]]$labels$title <- title
}
```

```{r}
for (i in 1:length(efs_plots)){
  fileName <- paste(names(efs_plots)[i], "_EFS_2017.03.30.pdf", sep = "")
  # pdf(file = fileName, height = 5, width = 6)
  print(efs_plots[[i]])
  # dev.off()
}
```


#Mutation EFS Analysis 
```{r}
names(mutationsDataFrames)
```

###Survival Analysis
```{r}
mutation_SA <- lapply(mutationsDataFrames, survivalAnalysis)
mutations_SA
```

###KM Curves
```{r}
efs_Mutplots <- mapply(efsCustomPlot, mutation_SA, colorsMut, SIMPLIFY = FALSE)
```

```{r}
summary(efs_Mutplots)
```

```{r}
for (mut in names(efs_Mutplots)){
  mutName <- paste0(toupper(substr(mut, 1,1)), substr(mut, 2,nchar(mut))) 
  title <- paste(c("Event Free Survival in \n Patients with", mutName, "Mutations" ), collapse = " ")
  print(title)
  efs_Mutplots[[mut]]$labels$title <- title
}
```

```{r}
for (i in 1:length(efs_Mutplots)){
  fileName <- paste(names(efs_Mutplots)[i], "_EFS_2017.03.30.pdf", sep = "")
#   pdf(file = fileName, height = 5, width = 6)
  print(efs_Mutplots[[i]])
#   dev.off()
}
```


#Two Group Comparisons for Mutations

```{r}
mutationDiff <- function(df1, df2, ID){
  #df1 - dataframe withonly the patients who have the mutation present
  #df2 - the larger dataframe to be subset to those without the mutation
  #ID is the column name matching patients IDs
  group1 <- df1
  group2 <- df2[ ! df2[,ID] %in% group1[,ID], ]
  mutation_status <- c(rep("present",nrow(group1)), rep("absent", nrow(group2)))

  comp_df <- rbind(group1,group2)
  comp_df <- cbind(mutation_status, comp_df)
  
  return(comp_df)
}
```



##Cohesins, myosins, and spliceosomes

### Create the dataframes with a column for pres/abs status of mutation. 
```{r}
twoGroupComparisons <- lapply(mutationsDataFrames, mutationDiff, df2=newClassification, ID="TARGET.USI")

```

### Check that this is the same as seen in single line EFS curves
```{r}
sapply(twoGroupComparisons, function(x) length(which(x$mutation_status == "present")))
```

### Suvival Analysis
```{r}
comparisons_SA <- lapply(twoGroupComparisons, SurvObjects, group = "mutation_status", degFree = 1, type = "EFS")

comparisons_SA
```


### KM Plots 
```{r}
fits <- lapply(comparisons_SA, function(l) l[[1]]) #selects only the first items in the list, which is the fit
colors<- list(cohesin=c("dodgerblue4","royalblue1"), spliceosome=c("dodgerblue4","darkturquoise"), myosin=c("dodgerblue4", "springgreen3"))
```


```{r}
comparisons_Plots <- mapply(efsCustomPlot, fits, colors, SIMPLIFY = FALSE)
```

```{r}
for (plot in names(comparisons_Plots)){
  mutName <- paste0(toupper(substr(plot, 1,1)), substr(plot, 2,nchar(plot))) 
  title <- paste(c("Event Free Survival in \n Patients with", mutName, " Fusions" ), collapse = " ")
  print(title)
  comparisons_Plots[[plot]]$labels$title <- title
}
```


```{r}
for (i in 1:length(comparisons_Plots)){
  fileName <- paste(names(comparisons_Plots)[i], "_EFS_2017.03.30.pdf", sep = "")
  # pdf(file = fileName, height = 5, width = 6)
  print(comparisons_Plots[[i]])
  # dev.off()
}
```


#cKit Mutations efs

```{r}

group1 <- newClassification[which(grepl("RUNX1-RUNX1T1 \\+ KIT", newClassification$Reason) | grepl("CBFB-MYH11 \\+ KIT", newClassification$Reason)), ]

CBF_KIT <- rep("CBF and cKIT", nrow(group1))

group1 <- cbind(CBF_KIT, group1)
```

```{r}
group2 <- newClassification[which(grepl("RUNX1-RUNX1T1", newClassification$Reason) | grepl("CBFB-MYH11", newClassification$Reason)), ]

group2 <- group2[ ! group2$TARGET.USI %in% group1$TARGET.USI, ]

CBF_KIT <- rep("CBF Alone", nrow(group2))

group2 <- cbind(CBF_KIT, group2)
```

```{r}
dim(group1)
dim(group2)
```


```{r}
KIT_df <- rbind(group1,group2)

head(KIT_df[,c(1,9:12)])
```

### Survival analysis
```{r message=FALSE, warning=FALSE}
KIT_SA <- SurvObjects(KIT_df, "CBF_KIT", 1, "EFS")

KIT_SA
```


### KM plots

```{r message=FALSE, warning=FALSE}
# KIT <- SurvivalPlot(KIT_SA[[1]], "Mutation Status", "Year", c("dodgerblue4","darkturquoise"))
KIT <- efsCustomPlot(KIT_SA[[1]],  c("dodgerblue4","darkturquoise"))
```

```{r}
KIT$labels$title <- "Event Free Survival in \n Patients with CBF and cKIT Mutations"

KIT
```


```{r}
# pdf(file = "CBF_KIT_2017.03.31.pdf", height = 5, width = 8)
# KIT
# dev.off()
```




